---
output: github_document
always_allow_html: true
editor_options: 
  markdown: 
    wrap: 72
  chunk_output_type: console
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%",
  message = FALSE,
  warning = FALSE,
  fig.retina = 2,
  fig.align = 'center'
)
```

# solidwastekampala

<!-- badges: start -->

[![License: CC BY
4.0](https://img.shields.io/badge/License-CC_BY_4.0-lightgrey.svg)](https://creativecommons.org/licenses/by/4.0/)
[![R-CMD-check](https://github.com/openwashdata/solidwastekampala/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/openwashdata/solidwastekampala/actions/workflows/R-CMD-check.yaml)
[![DOI](https://zenodo.org/badge/DOI/10.5281/zenodo.21519796.svg)](https://zenodo.org/doi/10.5281/zenodo.21519796)
<!-- badges: end -->

The solidwastekampala package provides the household-level data behind
the paper "Quantity and composition of domestic solid waste in Kampala
City as influenced by socioeconomic factors" by Katukiza et al. (2026),
published in *Frontiers in Environmental Science*
([doi:10.3389/fenvs.2026.1889921](https://doi.org/10.3389/fenvs.2026.1889921)).
Over a seven-day measurement campaign, the study
team weighed the domestic solid waste generated by 103 households in
three Kampala parishes selected to represent different income levels:
Bwaise I (low), Bukoto I (middle), and Ggaba (high). Each record
combines the measured mass of ten waste categories with derived
per-household metrics and the socio-economic characteristics of the
household and its dwelling. The package contains one analysis-ready
dataset together with a full variable dictionary.

## Associated publication

If you use these data, please cite the paper they belong to:

> Katukiza, A. Y., Niwagaba, C. B., Feni, I., Namagembe, S., Semiyaga,
> S., Batte, A., & Manga, M. (2026). Quantity and composition of
> domestic solid waste in Kampala City as influenced by socioeconomic
> factors. *Frontiers in Environmental Science*, 14.
> <https://doi.org/10.3389/fenvs.2026.1889921>

``` bibtex
@article{10.3389/fenvs.2026.1889921,
  author  = {Katukiza, Alex Y. and Niwagaba, Charles B. and Feni, Ivan and
             Namagembe, Shaluwa and Semiyaga, Swaib and Batte, Abubakar and
             Manga, Musa},
  title   = {Quantity and composition of domestic solid waste in Kampala
             City as influenced by socioeconomic factors},
  journal = {Frontiers in Environmental Science},
  volume  = {14},
  year    = {2026},
  doi     = {10.3389/fenvs.2026.1889921},
  url     = {https://www.frontiersin.org/journals/environmental-science/articles/10.3389/fenvs.2026.1889921},
  issn    = {2296-665X}
}
```

Please also cite the data package itself (see [Citation](#citation)).

## Installation

You can install the development version of solidwastekampala from
[GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("openwashdata/solidwastekampala")
```

```{r}
## Run the following code in console if you don't have the packages
## install.packages(c("dplyr", "knitr", "readr", "stringr", "gt", "kableExtra"))
library(dplyr)
library(knitr)
library(readr)
library(stringr)
library(gt)
library(kableExtra)
```

Alternatively, you can download the individual datasets as a CSV or XLSX
file from the table below.

1.  Click Download CSV. A window opens that displays the CSV in
    your browser.
2.  Right-click anywhere inside the window and select "Save Page As...".
3.  Save the file in a folder of your choice.

```{r, echo=FALSE, message=FALSE, warning=FALSE}

extdata_path <- "https://github.com/openwashdata/solidwastekampala/raw/main/inst/extdata/"

read_csv("data-raw/dictionary.csv") |> 
  distinct(file_name) |> 
  dplyr::mutate(file_name = str_remove(file_name, ".rda")) |> 
  dplyr::rename(dataset = file_name) |> 
  mutate(
    CSV = paste0("[Download CSV](", extdata_path, dataset, ".csv)"),
    XLSX = paste0("[Download XLSX](", extdata_path, dataset, ".xlsx)")
  ) |> 
  knitr::kable()

```

## Data

The package provides access to one dataset with the waste measurements
and socio-economic characteristics of the 103 analyzed households.

```{r}
library(solidwastekampala)
```

### solidwastekampala

The dataset `solidwastekampala` contains the mass of ten waste
categories collected from each household over a seven-day period, the
derived per-household metrics (total, per-day, per-capita, volume, and
density), and the socio-economic characteristics of the household and
its dwelling. It has
`r nrow(solidwastekampala)` observations and `r ncol(solidwastekampala)`
variables.

```{r}
solidwastekampala |> 
  head(3) |> 
  gt::gt() |>
  gt::as_raw_html()
```

For an overview of the variable names, see the following table.

```{r echo=FALSE, message=FALSE, warning=FALSE}
readr::read_csv("data-raw/dictionary.csv") |>
  dplyr::filter(file_name == "solidwastekampala.rda") |>
  dplyr::select(variable_name:description) |> 
  knitr::kable() |> 
  kableExtra::kable_styling("striped") |> 
  kableExtra::scroll_box(height = "200px")
```

## Example

The example reproduces the headline figures of the paper: mean
per-capita waste generation and mean waste density by income level.

```{r}
library(solidwastekampala)
library(dplyr)
library(ggplot2)

solidwastekampala |>
  group_by(income_level) |>
  summarise(
    households = n(),
    waste_kg_capita_day = round(mean(waste_per_capita_kg), 2),
    density_kg_m3 = round(mean(density_kg_l) * 1000)
  )
```

The figure below shows the households behind those numbers, one dot per
household. Waste per person rises with the income level of the study
area: the median goes from 0.32 kg per person per day in Bwaise I (low)
to 0.57 kg in Bukoto I (middle) and 0.93 kg in Ggaba (high), where the
spread across households is also widest. The article [Waste per person
by income](https://openwashdata.github.io/solidwastekampala/articles/waste-per-person.html)
shows an interactive version of this chart.

```{r dotplot-per-capita, fig.height = 5}
households <- solidwastekampala |>
  mutate(
    title = paste0(
      tools::toTitleCase(as.character(income_level)), " income (",
      parish, "), ", n(), " households"
    ),
    .by = income_level
  ) |>
  mutate(title = reorder(title, as.integer(income_level)))

medians <- households |>
  summarise(median = median(waste_per_capita_kg), .by = title)

ggplot(households, aes(x = waste_per_capita_kg)) +
  geom_dotplot(
    binwidth = 0.035, method = "histodot", stackdir = "center",
    stackratio = 1.1, fill = "#8a3a8a", colour = "white", alpha = 0.75
  ) +
  geom_vline(
    data = medians, aes(xintercept = median),
    colour = "#b04208", linewidth = 1
  ) +
  geom_text(
    data = medians,
    aes(
      x = median, y = -0.5,
      label = paste("median", round(median, 2), "kg")
    ),
    colour = "#b04208", fontface = "bold", hjust = -0.05, vjust = 0,
    size = 3.6
  ) +
  facet_wrap(~ title, ncol = 1) +
  scale_x_continuous(limits = c(0, 2), breaks = seq(0, 2, 0.5)) +
  scale_y_continuous(NULL, breaks = NULL) +
  labs(
    x = "kg of waste per person per day",
    title = "Waste per person rises with income",
    subtitle = "One dot per household, with the median of each income level"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    strip.text = element_text(face = "bold", hjust = 0, size = 11),
    panel.grid.minor = element_blank(),
    plot.title = element_text(face = "bold"),
    plot.title.position = "plot"
  )
```

## License

Data are available as
[CC-BY](https://github.com/openwashdata/solidwastekampala/blob/main/LICENSE.md).

## Citation

Please cite this package using:

```{r}
citation("solidwastekampala")
```
