---
output: github_document
always_allow_html: true
editor_options: 
  markdown: 
    wrap: 72
  chunk_output_type: console
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%",
  message = FALSE,
  warning = FALSE,
  fig.retina = 2,
  fig.align = 'center'
)
```

# solidwastekampala

<!-- badges: start -->

[![License: CC BY
4.0](https://img.shields.io/badge/License-CC_BY_4.0-lightgrey.svg)](https://creativecommons.org/licenses/by/4.0/)
[![R-CMD-check](https://github.com/openwashdata/solidwastekampala/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/openwashdata/solidwastekampala/actions/workflows/R-CMD-check.yaml)
[![DOI](https://zenodo.org/badge/DOI/10.5281/zenodo.21519797.svg)](https://doi.org/10.5281/zenodo.21519797)
<!-- badges: end -->

The solidwastekampala package provides the household-level data behind
the manuscript "Quantity and Composition of Domestic Solid Waste in
Kampala City as Influenced by Socio-Economic Factors" (Katukiza et al.,
Makerere University). Over a seven-day measurement campaign, the study
team weighed the domestic solid waste generated by 103 households in
three Kampala parishes selected to represent different income levels:
Bwaise I (low), Bukoto I (middle), and Ggaba (high). Each record
combines the measured mass of ten waste categories with derived
per-household metrics and the socio-economic characteristics of the
household and its dwelling. The package contains one analysis-ready
dataset together with a full variable dictionary.

## Installation

You can install the development version of solidwastekampala from
[GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("openwashdata/solidwastekampala")
```

```{r}
## Run the following code in console if you don't have the packages
## install.packages(c("dplyr", "knitr", "readr", "stringr", "gt", "kableExtra"))
library(dplyr)
library(knitr)
library(readr)
library(stringr)
library(gt)
library(kableExtra)
```

Alternatively, you can download the individual datasets as a CSV or XLSX
file from the table below.

1.  Click Download CSV. A window opens that displays the CSV in
    your browser.
2.  Right-click anywhere inside the window and select "Save Page As...".
3.  Save the file in a folder of your choice.

```{r, echo=FALSE, message=FALSE, warning=FALSE}

extdata_path <- "https://github.com/openwashdata/solidwastekampala/raw/main/inst/extdata/"

read_csv("data-raw/dictionary.csv") |> 
  distinct(file_name) |> 
  dplyr::mutate(file_name = str_remove(file_name, ".rda")) |> 
  dplyr::rename(dataset = file_name) |> 
  mutate(
    CSV = paste0("[Download CSV](", extdata_path, dataset, ".csv)"),
    XLSX = paste0("[Download XLSX](", extdata_path, dataset, ".xlsx)")
  ) |> 
  knitr::kable()

```

## Data

The package provides access to one dataset with the waste measurements
and socio-economic characteristics of the 103 analyzed households.

```{r}
library(solidwastekampala)
```

### solidwastekampala

The dataset `solidwastekampala` contains the mass of ten waste
categories collected from each household over a seven-day period, the
derived per-household metrics (total, per-day, per-capita, volume, and
density), and the socio-economic characteristics of the household and
its dwelling. It has
`r nrow(solidwastekampala)` observations and `r ncol(solidwastekampala)`
variables.

```{r}
solidwastekampala |> 
  head(3) |> 
  gt::gt() |>
  gt::as_raw_html()
```

For an overview of the variable names, see the following table.

```{r echo=FALSE, message=FALSE, warning=FALSE}
readr::read_csv("data-raw/dictionary.csv") |>
  dplyr::filter(file_name == "solidwastekampala.rda") |>
  dplyr::select(variable_name:description) |> 
  knitr::kable() |> 
  kableExtra::kable_styling("striped") |> 
  kableExtra::scroll_box(height = "200px")
```

## Example

The example reproduces the headline figures of the manuscript: mean
per-capita waste generation and mean waste density by income level.

```{r}
library(solidwastekampala)
library(dplyr)
library(ggplot2)

solidwastekampala |>
  group_by(income_level) |>
  summarise(
    households = n(),
    waste_kg_capita_day = round(mean(waste_per_capita_kg), 2),
    density_kg_m3 = round(mean(density_kg_l) * 1000)
  )
```

The figure `boxplot-per-capita` below shows the distribution behind
those means: per-capita waste generation rises with the income level of
the study area, from 0.43 kg per person per day in Bwaise I (low) to
0.98 kg in Ggaba (high), where the spread across households is also
widest.

```{r boxplot-per-capita}
ggplot(solidwastekampala,
       aes(x = income_level, y = waste_per_capita_kg)) +
  geom_boxplot() +
  labs(
    x = "Income level of study area",
    y = "Waste generation (kg per person per day)",
    title = "Per-capita waste generation by income level"
  ) +
  theme_minimal()
```

## License

Data are available as
[CC-BY](https://github.com/openwashdata/solidwastekampala/blob/main/LICENSE.md).

## Citation

Please cite this package using:

```{r}
citation("solidwastekampala")
```
